---
title: "2024-11-18 training part 1"
author: "Anna"
date: "2024-11-18"
output:
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#course part 1 - laboratory
we selected 20 isolates
- 6 isolates are **klebsiella sp.**
- 6 isolates are ***E. coli sp.***
- 8 isolates are ***salmonella sp.***

# Installation of pakages

```{r}
```

```{r}
install.packages("here")
install.packages("DiagrammeR")
```
ctrl+alt+I (make a cell-shotucut on the keyboard)

```{r}
angry_or_not <- "not"
```

give content of the object (contsiner)
```{r}
angry_or_not
```

Anna, are you angry or not? 'r print (angry_not)'. Something else


# working with your data


```{r}
here::here()
```

```{r}
path_data <- here::here("data/2024-09-25_INIKA_SAMPLING_ANALYSIS_HUMAN - 2024-09-25_INIKA_SAMPLING_ANALYSIS_HUMAN.csv")
path_data
```

```{r}
human_data <-
  readr::read_csv(path_data)
```
# Day 3

## Packages into memory
```{r}
library(here)
library(tidyverse)
```

## Getting help - and what functions do

```{r}
?library()
help("library")
```

## we get out data
```{r}
path_data <- here::here("data/2024-09-25_INIKA_SAMPLING_ANALYSIS_HUMAN - 2024-09-25_INIKA_SAMPLING_ANALYSIS_HUMAN.csv")
path_data
```

```{r}
human_data <-
  readr::read_csv(path_data)
```
## Exprolation of a data frame

- Names of columns
```{r}
colnames(human_data)
```

- getting the columns names into a vector for pasting and manipulations
- I can use a function within a function !
```{r}
dput(colnames(human_data))
```
- Explanation: c() is to make vectors

```{r}
test <- c("one", "2", 3)
test
test[3]
typeof(test[3])
typeof(test)

class(test)
```
All the      elements of a vector are of the same 'type' (text, double = Real, interpreter, date ...)

```{r}
test2 <-c(1,2,3)
typeof(test2)
class(test2)
test2*test2
```

```{r}
test2 * c(1,2)
```



- overview of data and data types in your data

```{r}
glimpse(human_data)
str(human_data)
```

 - renaming columns names (easier to use)
 
```{r}
original_names <- colnames(human_data)
original_names
```
 
```{r}
library(tidyverse)
new_names <- str_replace_all(original_names, " ", "_")
new_names
new_names[88]
```
 

 - regex pattern
```{r}
new_names2 <-str_replace_all(new_names, "[?/,;.*()-]", "_") %>%
  str_remove_all("(^_*)|(_*$)")


new_names2
new_names2[88]
```

```{r}
colnames(human_data) <- new_names2
```

 - selecting columns
 
```{r}
human_data %>%
  select (INIKA_OH_TZ_ID, Ages_years, gender) %>%
  head()
```
 
```{r}
human_data_selection <- 
  human_data %>%
  select(INIKA_OH_TZ_ID, Age__yrs, Gender, Enter_a_date, Region, District, 
         Specify_if_other_district, Sample, Season, Origin_of_sample, 
         Which_class_grade_are_you, 
         Who_is_your_caretaker, 
         If_others__mention, 
         What_is_your_occupation_and_or_of_your_caretaker, 
         Have_you_ever_heard_about_AMR, If_yes__how_did_you_get_this_information, 
         Have_you_or_your_children_used_any_antibiotics_at_any_time, 
         If_yes__where_did_you_get_these_drugs_from,
         If_it_was_drug_sellers_or_pharmacy__did_you_have_a_prescription_from_the_doctor_prescriber,
         GPS_coordinates_latitude, GPS_coordinates_longitude) 

human_data_selection %>% View()
```

```{r}
human_data <-
  readr::read_csv(path_data)
```


```{r}
new_columns <- 
  colnames(human_data) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all( "[?/,;.*()-]", "_") %>%
  str_remove_all("(^_*)|(_*$)")
```

```{r}
colnames(human_data) <- new_columns
```

```{r}
human_data_selection <- 
  human_data %>%
  select(INIKA_OH_TZ_ID, Age__yrs, Gender, Enter_a_date, Region, District, 
         Specify_if_other_district, Sample, Season, Origin_of_sample, 
         Which_class_grade_are_you, 
         Who_is_your_caretaker, 
         If_others__mention, 
         What_is_your_occupation_and_or_of_your_caretaker, 
         Have_you_ever_heard_about_AMR, If_yes__how_did_you_get_this_information, 
         Have_you_or_your_children_used_any_antibiotics_at_any_time, 
         If_yes__where_did_you_get_these_drugs_from,
         If_it_was_drug_sellers_or_pharmacy__did_you_have_a_prescription_from_the_doctor_prescriber,
         GPS_coordinates_latitude, GPS_coordinates_longitude) 

human_data_selection %>% View()
human_data_selection %>% head()
```

```{r}
glimpse(human_data_selection)
str(human_data_selection)
```

 # HERE
  - something does not make sense - empty column
  
```{r}
human_data_selection <-
  human_data_selection %>%
  select(-Specify_if_other_district) %>%
  mutate_at(vars(Age__yrs) , as.integer) %>%
  mutate_if(is.character, factor) %>%
  mutate_at(vars(INIKA_OH_TZ_ID) , as.character)
```

```{r}
str(human_data_selection)
```

```{r}
levels(test&regions)
```

- summary
```{r}
summary(
  human_data_selection
)
```

 - looking at the content of the single column
 
  - different columns has different values
  
   - changing the types of column so they can be processed correctly
 
 
 # filter out data/values
 
```{r}
human_data_selection %>%
  filter(INIKA_OH_TZ_ID == "238")
```

```{r}
human_data_selection %>%
  filter(INIKA_OH_TZ_ID != "238") %>%
  filter(is.na(Gender))
  #summary()

#%>%
# view()
```

 




